---
title: "Train Stops Monthly Ridership"
output: html_document
date: "2023-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Datasets
```{r}
# Load required packages
library(tidyverse)

# Load the ridership data for trains
train_ridership <- read.csv("CTA Train Ridership.csv")
head(train_ridership)

# Load the frequency data for trains
train_frequency <- readxl::read_xlsx('Train frequencies.xlsx')
head(train_frequency)
```

### Joining frequencies and ridership
```{r}
# Merge datasets
train_ridership_frequency <- merge(train_ridership, train_frequency, by.x = c('stationname', 'daytype'), by.y = c('station', 'day'))
head(train_ridership_frequency)

# Remove data frames not needed anymore
rm(train_ridership)
rm(train_frequency)

# Checking data types of each column
str(train_ridership_frequency)
```

### Converting type of the date column
```{r}
# Conversion of date column to correct format
train_ridership_frequency <- mutate(train_ridership_frequency, date = as.Date(date, format = '%m/%d/%Y'))

# Check data type
str(train_ridership_frequency)
```

### Filter data from 2013 on, to align with property data
```{r}
# Data for only 2013 - 2023
train_ridership_frequency <- filter(train_ridership_frequency, date > as.Date('01/01/2012', format = '%m/%d/%Y'))

# View data
head(train_ridership_frequency)
str(train_ridership_frequency)
```
### Summarize the results for each station by month
```{r}
annual_train <- train_ridership_frequency%>%
  
  # Drop 'daytype'
  select(-daytype) %>%
  
  # Create columns for year and month
  mutate(year = lubridate::year(date)) %>%
  
  # Drop the 'date' column
  select(-date) %>%
  
  # Group by station_id, year, and month
  group_by(station_id, year) %>%
  
  # Calculate the mean of num_visits and rides
  summarise(total_frequency = sum(num_visits),
            total_ridership = sum(rides), .groups = 'drop')

unique(annual_train$station_id)
```



### Need to add locations of each train stop for spatial analysis (longitude and latitude)
```{r}
# Load the train locations
train_locations <- read.csv("coordinates of l stops.csv")
head(train_locations)

# Merge data frame with scores with locations data frame to get longitude and latitude
train_final <- left_join(annual_train, train_locations, by = c("station_id" = "MAP_ID"))
head(train_final)

# Remove data frames no longer need
rm(train_locations)
rm(annual_train)

# Droping irrelevant or duplicate columns
train_final$STOP_ID <- NULL
train_final$DIRECTION_ID <- NULL
train_final$STOP_NAME <- NULL
train_final$STATION_NAME <- NULL
train_final$STATION_DESCRIPTIVE_NAME <- NULL
train_final$ADA <- NULL
train_final$RED <- NULL
train_final$BLUE <- NULL
train_final$G <- NULL
train_final$BRN <- NULL
train_final$P <- NULL
train_final$Pexp <- NULL
train_final$Y <- NULL
train_final$Pnk <- NULL
train_final$O <- NULL

# View data
head(train_final)
dim(train_final)

# Remove duplicated rows
train_final <- train_final %>%
  distinct(station_id, year, .keep_all = TRUE)

# Verify data
head(train_final)
dim(train_final)
```

### Seperate the longitude and latitude column to two seperate columns
```{r}
# Load tidyr package
library(tidyr)

# Sperating location column in the train_final data set
train_final <- separate(train_final, Location, into = c("latitude", "longitude"), sep = ", ", convert = TRUE)
head(train_final)

# Remove parentheses from latitude and longitude columns that remain after column split
train_final$latitude <- gsub("\\(|\\)", "", train_final$latitude)
train_final$longitude <- gsub("\\(|\\)", "", train_final$longitude)

# Convert the resulting columns to numeric
train_final$latitude <- as.numeric(train_final$latitude)
train_final$longitude <- as.numeric(train_final$longitude)

# View dataset
head(train_final)
```


### Convert monthly_train into csv file
```{r}
csv_file_path <- "~/CTA Project/CTA Project/MSBA CTA Work Files/train annual data.csv"

write.csv(train_final, file = csv_file_path, row.names = FALSE)
```

