---
title: "Bus Stops Spatial Analysis"
output: html_document
date: "2023-11-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
```

### Load in the bus scores data
```{r}
bus_stops <- read.csv('bus data.csv')

#Convert longitude and latitude columns into numeric types
bus_stops$Latitude <- as.numeric(bus_stops$lattitude) 
bus_stops$Longitude <- as.numeric(bus_stops$longitude)
```

### Load in the property data and clean it up
```{r}
properties <- read.csv('property values 2013.csv')
names(properties)

#Clean our property data to only have columns we need.
prop_coord <- select(properties, c('prop_id', 'Longitude', 'Latitude', 'Sale.Date'))
head(prop_coord)

#Remove the columns
prop_coord <- na.omit(prop_coord)
head(prop_coord)

#Convert longitude and latitude columns into numeric types
prop_coord$Latitude <- as.numeric(prop_coord$Latitude) 
prop_coord$Longitude <- as.numeric(prop_coord$Longitude)
```

### Add month and year columns to the dataset
```{r}
#Convert Sale.Date into a date datatype
prop_coord <- mutate(prop_coord, date = as.Date(Sale.Date, format = '%m/%d/%Y'))

# Clean property data to have only relevant columns
prop_coord <- prop_coord %>%
  select(c("prop_id", "Longitude", "Latitude", "date")) %>%
  mutate(year = lubridate::year(date),
         month = lubridate::month(date, label = TRUE, abbr = FALSE)) %>%
  select(-date)
head(prop_coord)
```

### Convert the two data frames into spatial objects
```{r}

# Convert the data frames into spatial objects
#bus_sf <- st_as_sf(bus_stops[sample(nrow(bus_stops), 2000),], coords = c("longitude", #"lattitude"))
#property_sf <- st_as_sf(prop_coord[sample(nrow(prop_coord), 20000),], coords = c('Longitude', #'Latitude'))

# Free space by removing df's that are no longer needed
#rm(bus_stops)
#rm(properties)
#rm(prop_coord)

# Set CRS for spatial objects
#bus_sf <- st_set_crs(bus_sf, "+proj=longlat +datum=WGS84")

#property_sf <- st_set_crs(property_sf, "+proj=longlat +datum=WGS84")


#Set appropriate CRS (Allows us to measure distance in miles)
#bus_sf <- st_transform(bus_sf, crs = st_crs("+proj=utm +zone=16 +datum=WGS84 +units=m #+no_defs"))

#property_sf <- st_transform(property_sf, crs = st_crs("+proj=utm +zone=16 +datum=WGS84 +units=m #+no_defs"))
```

### Calculate the number of stops within a 0.5 mile radius for each property within a given month in a given year. Sum up the scores of the stops that is captured in radius to assign a property a transit score
```{r}
#library(dplyr)

# Define a function to calculate distances and average scores
#calculate_distances_and_scores <- function(property, bus) {
  
  # Buffer the property points to create a circle with a 0.5 mile radius
#  property_buffered <- st_buffer(st_as_sf(property), dist = 0.5 * 1609.34)  # 1 mile = 1609.34 #meters
  
  # Spatial join to identify train stations within the buffer for each property
#  result_df <- st_join(property_buffered, st_as_sf(bus))
  
  # Filter rows with the same year and month
#  result_df <- filter(result_df, year.x == year.y, month.x == month.y)
  
  # Group by property_id, year, and month, and calculate the average score
#  result_df %>%
#    group_by(prop_id, year.x, month.x) %>%
 #   summarise(total_score = sum(score),
#              bus_count = n())
#}

# Use the function to calculate distances and average scores
#result_df <- calculate_distances_and_scores(property_sf, bus_sf)

# View the resulting data frame
#head(result_df)
```
### Vizualize the data
```{r}
#ggplot() +
#  geom_sf(data = bus_sf, color = "red", size = 5, alpha = .1) +
#  geom_sf(data = property_sf, color = "blue", size = 2, alpha = .02) +
#  theme_minimal()
```

### Activate chunk below to create csv file for "Transit Score sample csv.RMD"
```{r}

# Load required packages
library(sf)
library(dplyr)

# Initialize an empty dataframe to store the results
combined_results <- data.frame()

# Repeat the process 100 times
for (i in 1:100) {
  # Convert the data frames into spatial objects
  bus_sf <- st_as_sf(bus_stops[sample(nrow(bus_stops), 2000),], coords = c("longitude", "lattitude"))
  property_sf <- st_as_sf(prop_coord[sample(nrow(prop_coord), 20000),], coords = c('Longitude', 'Latitude'))

  # Set CRS for spatial objects
  bus_sf <- st_set_crs(bus_sf, "+proj=longlat +datum=WGS84")
  property_sf <- st_set_crs(property_sf, "+proj=longlat +datum=WGS84")

  # Set appropriate CRS (Allows us to measure distance in miles)
  bus_sf <- st_transform(bus_sf, crs = st_crs("+proj=utm +zone=16 +datum=WGS84 +units=m +no_defs"))
  property_sf <- st_transform(property_sf, crs = st_crs("+proj=utm +zone=16 +datum=WGS84 +units=m +no_defs"))

  # Define a function to calculate distances and average scores
  calculate_distances_and_scores <- function(property, bus) {
    # Buffer the property points to create a circle with a 0.5 mile radius
    property_buffered <- st_buffer(st_as_sf(property), dist = 0.5 * 1609.34)  # 1 mile = 1609.34 meters
    
    # Spatial join to identify train stations within the buffer for each property
    result_df <- st_join(property_buffered, st_as_sf(bus))
    
    # Filter rows with the same year and month
    result_df <- filter(result_df, year.x == year.y, month.x == month.y)
    
    # Group by property_id, year, and month, and calculate the average score
    result_df %>%
      group_by(prop_id, year.x, month.x) %>%
      summarise(total_score = sum(norm_score),
                num_bus = n())
  }

  # Use the function to calculate distances and average scores
  result_df <- calculate_distances_and_scores(property_sf, bus_sf)

  # Bind the result to the combined_results dataframe
  combined_results <- rbind(combined_results, result_df)
}


# Save the combined results to a CSV file
write.csv(combined_results, "combined_results.csv", row.names = FALSE)
```
