---
title: "Train Stops Monthly Ridership"
output: html_document
date: "2023-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Datasets
```{r}
# Load required packages
library(tidyverse)

# Load the ridership data for trains
train_ridership <- read.csv("CTA Train Ridership.csv")
head(train_ridership)

# Load the frequency data for trains
train_frequency <- readxl::read_xlsx('Train frequencies.xlsx')
head(train_frequency)
```

### Joining frequencies and ridership
```{r}
# Merge datasets
train_ridership_frequency <- merge(train_ridership, train_frequency, by.x = c('stationname', 'daytype'), by.y = c('station', 'day'))
head(train_ridership_frequency)

# Remove data frames not needed anymore
rm(train_ridership)
rm(train_frequency)

# Checking data types of each column
str(train_ridership_frequency)
```

### Converting type of the date column
```{r}
# Conversion of date column to correct format
train_ridership_frequency <- mutate(train_ridership_frequency, date = as.Date(date, format = '%m/%d/%Y'))

# Check data type
str(train_ridership_frequency)
```

### Winsorizing the dataset
```{r}
winsorize <- function(x, low_perc = 0.05, high_perc = 0.95) {
  quantiles <- quantile(x, probs = c(low_perc, high_perc), na.rm = TRUE)
  x[x < quantiles[1]] <- quantiles[1]
  x[x > quantiles[2]] <- quantiles[2]
  x
}
# winsorize ridership column
train_ridership_frequency$rides <- winsorize(train_ridership_frequency$rides)
```

### Filter data from 2013 on, to align with property data
```{r}
# Data for only 2013 - 2023
train_ridership_frequency <- filter(train_ridership_frequency, date > as.Date('01/01/2013', format = '%m/%d/%Y'))

# View data
head(train_ridership_frequency)
str(train_ridership_frequency)
```
### Summarize the results for each station by month
```{r}
monthly_train <- train_ridership_frequency%>%
  
  # Drop 'daytype'
  select(-daytype) %>%
  
  # Create columns for year and month
  mutate(year = lubridate::year(date),
         month = lubridate::month(date, label = TRUE, abbr = FALSE)) %>%
  
  # Drop the 'date' column
  select(-date) %>%
  
  # Group by station_id, year, and month
  group_by(station_id, year, month) %>%
  
  # Calculate the mean of num_visits and rides
  summarise(avg_frequency = mean(num_visits),
            avg_rides = mean(rides), .groups = 'drop')

head(monthly_train)
```

### Assigning a normalized relative score to every train stop
```{r}
# Calculating the transit scores
train_scores <- mutate(monthly_train, norm_rides = avg_rides / max(avg_rides) * 100,
                                      norm_frequency = avg_frequency / max(avg_frequency) * 100,
                                      score = ((norm_rides + 1) * (norm_frequency + 1)) ^ (1/3),                                       norm_score = score / max(score) * 100)

# View data
head(train_scores)

# Remove unused data frame
rm(monthly_train)
```
### Distribution of Scores
```{r}
# Summary of the scores
summary(train_scores$score)
summary(train_scores$norm_score)

# Load ggplot2
library(ggplot2)

# Create histogram of normalized transit scores
ggplot(train_scores, aes(x = norm_score)) +
  geom_histogram(bins = 50)
```
### Need to add locations of each train stop for spatial analysis (longitude and latitude)
```{r}
# Load the train locations
train_locations <- read.csv("coordinates of l stops.csv")
head(train_locations)

# Merge data frame with scores with locations data frame to get longitude and latitude
train_final <- left_join(train_scores, train_locations, by = c("station_id" = "MAP_ID"))
head(train_final)

# Remove data frames no longer need
rm(train_locations)
rm(train_scores)

# Droping irrelevant or duplicate columns
train_final$STOP_ID <- NULL
train_final$DIRECTION_ID <- NULL
train_final$STOP_NAME <- NULL
train_final$STATION_NAME <- NULL
train_final$STATION_DESCRIPTIVE_NAME <- NULL
train_final$ADA <- NULL
train_final$RED <- NULL
train_final$BLUE <- NULL
train_final$G <- NULL
train_final$BRN <- NULL
train_final$P <- NULL
train_final$Pexp <- NULL
train_final$Y <- NULL
train_final$Pnk <- NULL
train_final$O <- NULL

# View data
head(train_final)
dim(train_final)

# Remove duplicated rows
train_final <- train_final %>%
  distinct(station_id, year, month, .keep_all = TRUE)

# Verify data
head(train_final)
dim(train_final)
```

### Seperate the longitude and latitude column to two seperate columns
```{r}
# Load tidyr package
library(tidyr)

# Sperating location column in the train_final data set
train_final <- separate(train_final, Location, into = c("latitude", "longitude"), sep = ", ", convert = TRUE)
head(train_final)

# Remove parentheses from latitude and longitude columns that remain after column split
train_final$latitude <- gsub("\\(|\\)", "", train_final$latitude)
train_final$longitude <- gsub("\\(|\\)", "", train_final$longitude)

# Convert the resulting columns to numeric
train_final$latitude <- as.numeric(train_final$latitude)
train_final$longitude <- as.numeric(train_final$longitude)

# View dataset
head(train_final)
```


### Convert monthly_train into csv file
```{r}
csv_file_path <- "/Users/joel3/OneDrive/Documents/CTA Project/Relative Scores/train monthly scores.csv"

write.csv(train_final, file = csv_file_path, row.names = FALSE)
```

