---
title: "Monthly Bus Relative Scores"
output: html_document
date: "2023-11-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
```
### Load in ridership data
```{r}
# Load in bus route ridership dataset
bus_ridership <- read_csv('CTA Bus Ridership.csv')
head(bus_ridership)

# Convert date into proper data type
bus_ridership <- mutate(bus_ridership, date = as.Date(date, format = '%m/%d/%Y'))
str(bus_ridership)

# filter for dates after 2013
bus_rides_2013 <- filter(bus_ridership, date > as.Date('2013-01-01') & date < as.Date('2020-01-01'))
head(bus_rides_2013)
```
### Load in the frequency data
```{r}
# Load in the bus route frequency data
ride_visits <- readxl::read_xlsx('MSBA CTA.xlsx')
head(ride_visits)

# Remove unwanted column
ride_visits <- ride_visits[-4]
head(ride_visits)
```
### Join the two datasets
```{r}
#Join the frequency and ridership tables
bus_ride_freq <- left_join(bus_rides_2013, ride_visits, c('route' = 'route_number', 'daytype' = 'day'))
head(bus_ride_freq)

#Remove dataframes no longer needed to free up space
rm(bus_rides_2013)
rm(ride_visits)
rm(bus_ridership)

#Identify null values
null <- bus_ride_freq %>% filter(is.na(route_name)) %>% distinct(route)
print(null, n = 37)

#Get rid of null values
bus_route <- na.omit(bus_ride_freq)
head(bus_route)
```
### Convert Daily data into monthly data
```{r}
monthly_bus <- bus_route %>%
  
  # Drop 'daytype'
  select(-daytype) %>%
  
  # Create columns for year and month
  mutate(year = lubridate::year(date),
         month = lubridate::month(date, label = TRUE, abbr = FALSE)) %>%
  
  # Drop the 'date' column
  select(-date) %>%
  
  # Group by station_id, year, and month
  group_by(route, year, month) %>%
  
  # Calculate the mean of num_visits and rides
  summarise(avg_frequency = mean(num_of_visits),
            avg_rides = mean(rides), .groups = 'drop')

```
### Assign each observation a relative score
```{r}
# Calculate relative sores for each bus route using the following function
bus_scores <- mutate(monthly_bus, score = ((avg_rides + 1) * (avg_frequency + 1))^(1/3),
                    norm_score = score / max(score) * 100)

# Analyze distribution of relative scores
summary(bus_scores$score)
summary(bus_scores$norm_score)

# Visualize distribution of relative scores
ggplot(bus_scores, aes(x = norm_score)) +
  geom_histogram(bins = 50)
```
### Import bus stop data to convert route data to the stop level
```{r}
# Import bus stop data
bus_stops <- readxl::read_xlsx('Bus Stops.xlsx')
head(bus_stops)
```

### Join the bus stop data with the route data
```{r}
# Join the data
bus_stops <- left_join(bus_stops, bus_scores, c('route' = 'route'))

# Get rid of columns we no longer need to free up space
rm(bus_final)
rm(bus_ride_freq)
rm(monthly_bus)

# Get rid of nulls
bus_stops <- na.omit(bus_stops)
head(bus_stops)
```
### Visualize distribtuion of relative scores across all 
```{r}
ggplot(bus_stops, aes(x = norm_score)) +
  geom_histogram()
```
### Load in the dataset with coordinates for bus stops. Format the lattitude and longitude
```{r}
bus_coords<- read.csv('bus coords.csv')

# get rid of the parenthesis in the data
bus_coords$lattitude <- gsub("\\(|\\)", "", bus_coords$lattitude) 
bus_coords$longitude <- gsub("\\(|\\)", "", bus_coords$longitude)

# convert columns to the correct data types
bus_coords$lattitude <- as.numeric(bus_coords$lattitude) 
bus_coords$longitude <- as.numeric(bus_coords$longitude)
bus_coords$SYSTEMSTOP <- as.character(bus_coords$SYSTEMSTOP)

head(bus_coords)
```
### Join the coordinates with our data of relative scores
```{r}
# convert column to the correct data type
bus_stops$stop_id <- as.character(bus_stops$stop_id)

# Join the data we have for each stop with the coordinate of each stop
bus_spatial <- left_join(bus_stops, bus_coords, c('stop_id' = 'SYSTEMSTOP'))

# Remove null values
bus_spatial <- na.omit(bus_spatial)

# Only select columns of interest
bus_spatial <- bus_spatial %>%
  select(c(1:11)) %>%
  select(-2)

# Remove data frames no longer needed to free up space
rm(bus_stops)
rm(bus_coords)
```

### Convert bus_spatial into a csv file to continue work on new notebook
```{r}
csv_file_path <- "/Users/joel3/OneDrive/Documents/CTA Project/Relative Scores/bus data.csv"

write.csv(bus_spatial, file = csv_file_path, row.names = FALSE)
```

